---
title: "Metalearners vs ELTMLE for Heterogeneous Treatment Effect Estimation"
subtitle: "Adapted from Stata Blog (2025-10-22) — S-, T-, X-Learners (h2o + R) + ELTMLE (Stata) — Luque-Fernández et al."
author:
  - name: "Miguel Angel Luque-Fernández, MA, MSc, MPH, PhD — migariane"
    affiliations:
      - name: "Universidad de Granada (UGR)"
        department: "Dpto. Estadística e Investigación Operativa"
        url: "https://www.ugr.es"
      - name: "London School of Hygiene & Tropical Medicine (LSHTM)"
        department: "Honorary Associate Professor"
        url: "https://www.lshtm.ac.uk"
    orcid: "0000-0003-3552-4458"
    url: "https://migariane.github.io"
date: "2026-07-06"
format: html
engine: knitr
execute:
  eval: true
links:
  - text: "GitHub"
    url: "https://github.com/migariane"
  - text: "BioEstatR"
    url: "https://migariane.github.io/BioEstatR"
  - text: "Matemática Estadística con R"
    url: "https://migariane.github.io/MatematicaEstadisticaMedicinaR"
---

## Theoretical Framework

### From ATE to CATE: The Heterogeneity Problem

The **Average Treatment Effect** (ATE) summarizes the causal effect across the entire population:

$$\text{ATE} = E[Y(1) - Y(0)] = E[\tau_i]$$

The ATE is a single number. It answers: *"Does the treatment work, on average?"* But it masks crucial variation: the same treatment may benefit some subgroups, harm others, and have no effect on the rest. This is **treatment effect heterogeneity**.

The **Conditional Average Treatment Effect** (CATE) addresses this:

$$\tau(x) = E[Y(1) - Y(0) \mid X = x]$$

CATE answers: *"For a person with characteristics $x$, what is the expected effect?"* It is the workhorse of personalized medicine, targeted policy, and precision public health.

### The Fundamental Problem and the Role of Models

The **fundamental problem of causal inference** (Holland, 1986) is that we never observe both $Y_i(1)$ and $Y_i(0)$ for the same unit. The observed outcome is:

$$Y_i = T_i Y_i(1) + (1 - T_i) Y_i(0)$$

To estimate $\tau(x)$, we must **impute the missing counterfactual** using a model for the conditional expectations:

$$\mu_t(x) = E[Y \mid X = x, T = t], \quad t \in \{0, 1\}$$

The quality of any CATE estimator hinges on how well it estimates these **nuisance functions** $\mu_0(x)$ and $\mu_1(x)$ — plus, for doubly-robust methods, the **propensity score** $g(x) = P(T = 1 \mid X = x)$.

### Identification Assumptions

CATE is identified under three standard assumptions:

1. **Unconfoundedness** (conditional exchangeability): $(Y(0), Y(1)) \perp T \mid X$. All confounders are measured and included in $X$. This is the central — and untestable — assumption in observational causal inference.

2. **Overlap** (positivity): $0 < P(T = 1 \mid X = x) < 1$ for all $x$. There are no subgroups where everyone (or no one) receives treatment. Violations lead to extrapolation — predicting outcomes in regions with no data.

3. **SUTVA** (Stable Unit Treatment Value Assumption): each unit's potential outcomes depend only on its own treatment, not on others'. Rules out spillover, contagion, and general equilibrium effects.

### Why Not Just Regression?

A natural question: *why not just fit $E[Y \mid X, T]$ with interactions $T \times X$?*

Three problems:

1. **Model misspecification**: parametric forms (linear, logistic) rarely hold exactly. ML methods relax this but introduce regularization bias.
2. **Regularization bias**: penalized/ensemble methods shrink estimates toward zero. The bias in $\hat{\mu}_t(x)$ propagates to $\hat{\tau}(x)$.
3. **The wrong objective**: minimizing prediction error for $Y$ does not minimize error for $\tau = \mu_1 - \mu_0$. A model that predicts $Y$ well may estimate $\tau$ poorly.

Metalearners and TMLE address these problems through different strategies.

### CATE Estimation Strategies: A Taxonomy

| Approach | Examples | Key Idea |
|----------|----------|----------|
| **Outcome modeling** | S-learner, T-learner | Estimate $\mu_t(x)$ directly; take difference |
| **Pseudo-outcome regression** | X-learner, R-learner, DR-learner | Construct proxy outcomes that behave like $\tau_i$ |
| **Targeted smoothing** | TMLE, AIPW | Correct initial ML estimates with an efficient influence function step |
| **Tree/forest-based** | Causal trees, causal forest (grf) | Partition $X$-space to maximize treatment effect heterogeneity |

This document implements the first three approaches using GBM as base learner (metalearners) and the default Super Learner (ELTMLE/tmle).

### Software Frameworks

#### H2O.ai (R/Python/Java)

[H2O](https://h2o.ai) is an open-source, distributed ML platform. Its R interface (`h2o` package, [docs.h2o.ai](https://docs.h2o.ai)) provides:

- **In-memory cluster**: Java backend; `h2o.init()` starts a local cluster. Data lives in H2O frames, not R memory.
- **AutoML + GBM, RF, GLM, Deep Learning**: gradient boosting (`h2o.gbm`) used here as base learner.
- **Scalability**: designed for N > 10M. Handles factor/categorical encoding automatically.
- **Integration**: data flows R → H2O via `as.h2o()`, predictions back via `as.data.table()`.

```r
# Core workflow
h2o.init()                     # Start local cluster
hdf <- as.h2o(df, "frame_id")  # Upload to H2O
m <- h2o.gbm(y, x, hdf)        # Train
p <- h2o.predict(m, newdata)   # Predict
```

#### ELTMLE (Stata)

`eltmle` ([SSC](https://ideas.repec.org/c/boc/bocode/s458337.html), [GitHub](https://github.com/migariane/eltmle)) is a Stata/R wrapper implementing TMLE with Super Learner. Syntax:

```stata
ssc install eltmle
eltmle Y X Z [, tmle tmlebgam tmleglsrf bal elements cvtmle cvtmlebgam cvtmleglsrf cvfolds(#) seed(#)]
```

where **Y** = binary/continuous outcome, **X** = binary treatment (0/1), **Z** = covariates.

| Option | Learners added to default* | Use case |
|--------|---------------------------|----------|
| `tmle` (default) | SL: stepwise + GLM + GLM-poly-int | N < 50K, baseline |
| `tmlebgam` | + Bayes GLM + GAM | Non-linear effects |
| `tmleglsrf` | + Lasso + RF + GAM | Heterogeneous effects |

*Default Super Learner: stepwise selection, GLM, GLM with 2nd-order polynomials + two-way interactions. Cross-validated variants: `cvtmle`, `cvtmlebgam`, `cvtmleglsrf`.

### Metalearners

A **metalearner** decomposes CATE estimation into supervised regression tasks solved by off-the-shelf ML models (base learners). This document adapts the tutorial from the [Stata Blog (2025-10-22)](https://blog.stata.com/2025/10/22/heterogeneous-treatment-effect-estimation-with-s-t-and-x-learners-using-h2oml/), translating the original Stata/h2oml code to R/h2o and adding an ELTMLE comparison. Three metalearners using gradient boosting (GBM) via `h2o` are compared:

| Learner  | Strategy | Strength | Weakness |
|----------|----------|----------|----------|
| **S-learner** | Single model: $Y \sim X + T$ | Simple, efficient | May ignore $T$ in high-dimensional $X$ |
| **T-learner** | Two models: $\mu_0$, $\mu_1$ on separate groups | Treatment always considered | Less data per model |
| **X-learner** | Cross-estimation + imputed $\tau$ | Adapts to imbalance; uses full data | More complex, more steps |

### Data: Gerber, Green & Larimer (2008)

Field experiment on social pressure and voter turnout. Michigan households, August 2006 primary.
**Treatment:** self-and-neighbors mailer. **Outcome:** `voted` (binary). **N =** 229,461 (191,243 control + 38,218 treated).

---

## Setup {#setup}

Data loading and preprocessing: remove `haven_labelled` class (incompatible with `h2o`), create prior voting score (`totalvote`, 0--5), and upload to the H2O cluster.

```{r setup, message=FALSE, warning=FALSE}
library(h2o)
library(data.table)

h2o.init()

# Load: zap_labels() strips haven_labelled class (h2o incompatible)
df <- haven::read_dta("socialpressure.dta")
df <- haven::zap_labels(df)
df$totalvote <- df$g2000 + df$g2002 + df$p2000 + df$p2002 + df$p2004
df$voted <- as.factor(df$voted)
df$treatment <- as.factor(df$treatment)
setDT(df)

# Predictors: demographics + voting history + treatment
predictors <- c("gender", "g2000", "g2002", "p2000", "p2002", "p2004", "treatment", "age")

# Upload to H2O (local Java cluster)
hdf <- as.h2o(df, "social")
```

**Predictor variables**: gender, general election turnout (2000, 2002), primary turnout (2000, 2002, 2004), treatment arm, age.
**Outcome**: `voted` (voted in August 2006 primary).

::: {.callout-tip title="Reproducibility"}
**Data**: `socialpressure.dta` included in repository. Direct download:
[stata-press.com/data/r18/socialpressure.dta](https://www.stata-press.com/data/r18/socialpressure.dta)

**Requirements**: R ≥ 4.0, packages [`h2o`](https://docs.h2o.ai) (≥ 3.40), `data.table`, `haven`. Java 8+ required by H2O.
:::

---

## S-learner {#s-learner}

### Strategy

The S-learner ("Single") trains **a single model** that includes $T$ as just another predictor:

$$\hat{\mu}(x, t) = E[Y \mid X = x, T = t]$$

CATE is estimated via counterfactual difference:

$$\hat{\tau}_S(x) = \hat{\mu}(x, T = 1) - \hat{\mu}(x, T = 0)$$

For each unit, we predict $\hat{Y}$ twice: once forcing $T = 1$ and once forcing $T = 0$, using the same model.

```{r s-learner}
# Step 1: Train a single GBM with treatment as a predictor
s_model <- h2o.gbm(y = "voted", x = predictors,
  training_frame = hdf, seed = 19)

# Step 2: Build counterfactual frames (same data, T fixed)
hdf1 <- h2o.assign(hdf, "social1")
hdf1["treatment"] <- "1"   # Everyone treated
hdf0 <- h2o.assign(hdf, "social0")
hdf0["treatment"] <- "0"   # Everyone control

# Step 3: Predict under both scenarios
p1 <- as.data.table(h2o.predict(s_model, hdf1))  # Estimated Y(1)
p0 <- as.data.table(h2o.predict(s_model, hdf0))  # Estimated Y(0)

# Step 4: CATE = difference in predicted probabilities
df$catehat_S <- p1$p1 - p0$p1
summary(df$catehat_S)
df[, .(mean_cate = mean(catehat_S), n = .N), by = totalvote]
```

### Interpretation

The S-learner estimates a mean effect of **`r round(mean(df$catehat_S), 4)`** (difference in voting probability). Its limitation: when $X$ is high-dimensional, the model may "ignore" $T$ and produce $\hat{\tau} \approx 0$. The table shows how CATE varies with prior turnout (`totalvote`).

---

## T-learner {#t-learner}

### Strategy

The T-learner ("Two") trains **two separate models**:

$$\hat{\mu}_1(x) = E[Y \mid X = x, T = 1] \quad \text{(treated only)}$$
$$\hat{\mu}_0(x) = E[Y \mid X = x, T = 0] \quad \text{(controls only)}$$

CATE is obtained directly:

$$\hat{\tau}_T(x) = \hat{\mu}_1(x) - \hat{\mu}_0(x)$$

**Advantage**: $T$ is never diluted among predictors. **Disadvantage**: each model uses only a subset of data (efficiency loss).

```{r t-learner}
# Step 1: Split data by treatment group
treated_rows <- which(df$treatment == "1")   # 38,218 observations
control_rows <- which(df$treatment == "0")   # 191,243 observations

hdf0 <- as.h2o(df[control_rows, ], "social0")
hdf1 <- as.h2o(df[treated_rows, ], "social1")

# Step 2: Train two independent GBMs
m0 <- h2o.gbm(y = "voted", x = predictors,
  training_frame = hdf0, seed = 19)   # Control model
m1 <- h2o.gbm(y = "voted", x = predictors,
  training_frame = hdf1, seed = 19)   # Treated model

# Step 3: Predict on full dataset with each model
p0 <- as.data.table(h2o.predict(m0, hdf))   # μ̂₀ for all
p1 <- as.data.table(h2o.predict(m1, hdf))   # μ̂₁ for all

# Step 4: CATE as difference of predictions
df$catehat_T <- p1$p1 - p0$p1
summary(df$catehat_T)
df[, .(mean_cate = mean(catehat_T), n = .N), by = totalvote]
```

### Interpretation

Mean CATE (T-learner): **`r round(mean(df$catehat_T), 4)`**. By using separate models, the T-learner captures heterogeneity better when response surfaces $\mu_0$ and $\mu_1$ have very different shapes. Compared to the S-learner, estimates tend to be more variable (each model sees fewer data points).

---

## X-learner {#x-learner}

### Strategy

The X-learner (Künzel et al., 2019) leverages **all available information** in 5 steps:

1. **Estimate $\hat{\mu}_0$, $\hat{\mu}_1$** with separate models (same as T-learner)
2. **Impute individual effects** where unobserved:
   - For treated: $\tilde{D}_i^1 = Y_i(1) - \hat{\mu}_0(X_i)$ (observed outcome minus estimated control)
   - For controls: $\tilde{D}_i^0 = \hat{\mu}_1(X_i) - Y_i(0)$ (estimated treatment minus observed control)
3. **Model $\hat{\tau}_1(x) = E[\tilde{D}^1 \mid X = x]$** and $\hat{\tau}_0(x) = E[\tilde{D}^0 \mid X = x]$
4. **Weight by propensity score**: $\hat{\tau}_X(x) = g(x)\hat{\tau}_0(x) + (1 - g(x))\hat{\tau}_1(x)$
   where $g(x) = P(T = 1 \mid X = x)$

```{r x-learner}
# Step 1: Cross-prediction (control model → treated, and vice versa)
p0_on_1 <- as.data.table(h2o.predict(m0, hdf1))  # μ̂₀ predicted on treated
p1_on_0 <- as.data.table(h2o.predict(m1, hdf0))  # μ̂₁ predicted on controls

# Step 2: Impute individual effects (D̃)
D1 <- as.numeric(df$voted[treated_rows] == "1") - p0_on_1$p1  # Treated: Y(1) - μ̂₀
D0 <- p1_on_0$p1 - as.numeric(df$voted[control_rows] == "1")  # Controls: μ̂₁ - Y(0)

# Step 3: Estimate propensity score g(x) = P(T=1 | X)
ps_model <- h2o.gbm(y = "treatment", x = setdiff(predictors, "treatment"),
  training_frame = hdf, seed = 19)
gx <- as.data.table(h2o.predict(ps_model, hdf))$p1

# Step 4: Model τ̂₀(x) and τ̂₁(x) on imputed effects
dt1 <- cbind(D1 = D1, df[treated_rows, setdiff(predictors, "treatment"), with = FALSE])
dt0 <- cbind(D0 = D0, df[control_rows, setdiff(predictors, "treatment"), with = FALSE])

tau1_model <- h2o.gbm(y = "D1", x = setdiff(predictors, "treatment"),
  training_frame = as.h2o(dt1), seed = 19)
tau0_model <- h2o.gbm(y = "D0", x = setdiff(predictors, "treatment"),
  training_frame = as.h2o(dt0), seed = 19)

# Step 5: Predict and weight by propensity score
tau1 <- as.data.table(h2o.predict(tau1_model, hdf))$predict
tau0 <- as.data.table(h2o.predict(tau0_model, hdf))$predict

df$catehat_X <- gx * tau0 + (1 - gx) * tau1
summary(df$catehat_X)
df[, .(mean_cate = mean(catehat_X), n = .N), by = totalvote]
```

### Interpretation

Mean CATE (X-learner): **`r round(mean(df$catehat_X), 4)`**. The X-learner is especially useful with **imbalanced groups** (here the ratio is ~5:1 control/treated). By imputing effects in both directions, it uses data more efficiently. Propensity score weighting gives more weight to $\hat{\tau}_1$ where $g(x)$ is low (few treated units).

---

## ELTMLE: Ensemble Learning Targeted Maximum Likelihood Estimation {#eltmle}

### Theoretical Framework

**ELTMLE** ([`eltmle`](https://github.com/migariane/eltmle) for Stata; Luque-Fernández et al., 2018, 2023) combines two innovations:

1. **Targeted Maximum Likelihood Estimation (TMLE)**: a semiparametric estimator that corrects first-stage bias via a directed fluctuation (*targeting step*) on the initial estimate of $\mu_t(x)$. It is **doubly robust**: consistent if at least one of the two nuisance models — outcome $\mu_t(x)$ or propensity $g(x)$ — is correctly specified. It is also **asymptotically efficient** and **substitution-compatible** (respects parameter bounds, e.g., probabilities in [0,1]).

2. **Ensemble Learning (Super Learner)**: a cross-validated weighted combination of multiple candidate algorithms. `eltmle` offers:
   - `tmle` (default): Super Learner with SL: stepwise + GLM + GLM with polynomials and interactions
   - `tmlebgam`: adds Bayes GLM and GAM — suitable for non-linear effects
   - `tmleglsrf`: adds Lasso, Random Forest and GAM — suitable for heterogeneous effects

### Stata Code (Full Dataset)

```{bash, eval=FALSE}
#!/bin/bash
export PATH="/Applications/Stata/StataMP.app/Contents/MacOS:$PATH"

stata-mp -b -e '
  use socialpressure.dta, clear
  generate totalvote = g2000 + g2002 + p2000 + p2002 + p2004
  global covariates gender g2000 g2002 p2000 p2002 p2004 age
  eltmle voted treatment $covariates, tmle elements
  generate catehat_TMLE = _POM1 - _POM0
  summarize catehat_TMLE
  save tmle_full.dta, replace
'
```

::: {.callout-warning}
**N = 229,461**. The `elements` option retains `_POM1`, `_POM0`, and `_ps` variables, allowing computation of individual $\hat{\tau}_i$. Requires ~32 GB RAM for the Super Learner (10-fold CV × multiple algorithms). On machines with ≤ 16 GB, use `cvfolds(3)` or a 5% sample. The code block above is set to `eval=FALSE` due to memory constraints.
:::

### ELTMLE Results (tmle, 5% sample)

Run with `cvfolds(3)` on a 5% random sample (N = 11,473):

| Parameter | Estimate | SE | 95% CI | p-value |
|-----------|----------|-----|--------|---------|
| **ATE (TMLE)** | **0.057** | 0.011 | (0.034, 0.079) | <0.001 |
| Causal Risk Ratio | 1.19 | — | (1.12, 1.28) | — |
| Marginal OR | 1.30 | — | (1.17, 1.44) | — |

- **POM1** ($E[Y(1)]$): 0.347 — mean voting probability if everyone were treated
- **POM0** ($E[Y(0)]$): 0.291 — mean voting probability if everyone were control
- **Propensity score**: range [0.143, 0.183] — adequate overlap

**Interpretation**: the social pressure mailer increases voting probability by **5.7 percentage points** (95% CI: 3.4–7.9). The causal risk ratio of 1.19 indicates a 19% higher probability of voting under treatment.

```{bash, eval=FALSE}
# Reproducible execution (5% sample)
export PATH="/Applications/Stata/StataMP.app/Contents/MacOS:$PATH"
stata-mp -b -e '
  use socialpressure.dta, clear
  set seed 19
  sample 5
  generate totalvote = g2000 + g2002 + p2000 + p2002 + p2004
  global covariates gender g2000 g2002 p2000 p2002 p2004 age
  eltmle voted treatment $covariates, tmle elements cvfolds(3)
  generate catehat_TMLE = _POM1 - _POM0
  graph box catehat_TMLE, over(totalvote) yline(0) name(eltmle_plot, replace)
  graph export eltmle_boxplot.png, replace width(1200)
'
```

### ELTMLE vs Metalearners Comparison

| Dimension | S/T/X-learners (h2o GBM) | ELTMLE (Super Learner) |
|-----------|--------------------------|------------------------|
| **Base learner** | GBM (single algorithm) | GLM + GBM + RF + ... (ensemble) |
| **Robustness** | Depends on correct $\mu$ specification | Doubly robust ($\mu$ or $g$) |
| **Inference** | Point estimates only | Asymptotic CI and p-values |
| **Efficiency** | Not guaranteed | Asymptotically efficient |
| **CATE** | By strata (totalvote bins) | Continuous CATE in $X$ |
| **Software** | R (h2o) | Stata (eltmle) |
| **Computational load** | Moderate (~30s with GBM) | High (CV + ensemble × K folds) |

### When to Use Each Approach

- **S/T/X-learners**: rapid heterogeneity exploration, prototyping, teaching. The X-learner is the most robust of the three for observational data with imbalance.
- **ELTMLE (TMLE + Super Learner)**: formal inference, publication, ATE/ATT estimation with valid CIs. Requires more computation but provides frequentist guarantees under weaker conditions.

### ATE Summary by Method

:::: {.columns}
::: {.column width="60%"}

| Method | ATE | 95% CI | N |
|--------|-----|--------|---|
| S-learner (GBM) | `r round(mean(df$catehat_S), 3)` | — | 229,461 |
| T-learner (GBM) | `r round(mean(df$catehat_T), 3)` | — | 229,461 |
| X-learner (GBM) | `r round(mean(df$catehat_X), 3)` | — | 229,461 |
| **ELTMLE (tmle)** | **0.057** | (0.034, 0.079) | 11,473 |

:::
::: {.column width="40%"}

::: {.callout-note}
**Note**: S/T/X-learners do not provide CIs by construction — their ATEs are means of individual CATEs. ELTMLE is the only method with formal inference.

Agreement of ~0.07–0.08 across methods suggests the effect is robust.
:::

:::
::::

---

## Visual Comparison {#comparacion-visual}

### Metalearners (S, T, X)

CATE by prior turnout level (N = 229,461). Dashed red line: zero effect.

```{r plot-metals, fig.width=10, fig.height=4}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
boxplot(catehat_S ~ totalvote, data = df, main = "S-learner (GBM, N=229K)",
        ylab = "CATE", xlab = "Prior turnout (0-5)", las = 1,
        col = "steelblue")
abline(h = 0, lty = 2, col = "red", lwd = 2)
boxplot(catehat_T ~ totalvote, data = df, main = "T-learner (GBM, N=229K)",
        ylab = "CATE", xlab = "Prior turnout (0-5)", las = 1,
        col = "darkgreen")
abline(h = 0, lty = 2, col = "red", lwd = 2)
boxplot(catehat_X ~ totalvote, data = df, main = "X-learner (GBM, N=229K)",
        ylab = "CATE", xlab = "Prior turnout (0-5)", las = 1,
        col = "darkorange")
abline(h = 0, lty = 2, col = "red", lwd = 2)
```

```{r shutdown-h2o, include=FALSE}
h2o.shutdown(prompt = FALSE)
```

### ELTMLE (tmle, 5% sample)

Individual CATE obtained via the `elements` option: $\hat{\tau}_i =$ `_POM1` $-$ `_POM0` (N = 11,473).

```{r plot-eltmle, fig.width=5, fig.height=4}
tmle_df <- haven::read_dta("tmle_results.dta")
boxplot(catehat_TMLE ~ totalvote, data = tmle_df,
        main = "ELTMLE — tmle (N=11K, 5% sample)",
        ylab = "CATE", xlab = "Prior turnout (0-5)", las = 1,
        col = "#8B0000")
abline(h = 0, lty = 2, col = "red", lwd = 2)
```

### Reading the Charts

- Each boxplot shows the distribution of $\hat{\tau}(x)$ for units at a given level of prior turnout.
- **Expected pattern**: the treatment effect (social pressure mailer) is larger for those with medium-high voting history (3–4 prior elections) and smaller at the extremes (0 or 5).
- **Differences across learners**: the S-learner compresses effects toward zero; the T-learner shows more dispersion; the X-learner balances both tendencies.
- **ELTMLE**: the ELTMLE panel shows individual CATE as $\hat{\tau}_i =$ `_POM1` $-$ `_POM0` via the `elements` option. The pattern across `totalvote` is consistent with the metalearners — effect increasing up to 3–4 prior elections — though the scale is smaller when using a 5% sample.

---

## Discussion

### Which Metalearner to Choose?

| Criterion | Recommendation |
|-----------|----------------|
| Homogeneous or near-zero effect | **S-learner** — more efficient |
| Strong expected heterogeneity | **T-learner** — separate models capture different functional forms |
| Highly imbalanced groups | **X-learner** — imputes effects in both directions |
| Small sample / high dimensionality | **S-learner** — shares information across groups |
| Formal inference (CI, p-values) | **ELTMLE** — doubly robust, efficient |
| Publication / regulatory | **ELTMLE** — accepted standard in causal epidemiology |

### Limitations

1. **Untestable unconfoundedness**: we assume observed covariates block all confounding
2. **Overlap**: verify propensity score stays away from 0 and 1 across the support of $X$
3. **GBM as base learner**: sensitive to hyperparameters; an alternative is `grf::causal_forest` (R-learner, Athey, Tibshirani & Wager, 2019)
4. **Inference**: S/T/X-learner standard errors require bootstrap; ELTMLE provides asymptotic CIs by construction
5. **ELTMLE**: computationally expensive with N > 100K and multiple ensemble learners

---

## References

- Gerber, A. S., Green, D. P., & Larimer, C. W. (2008). Social pressure and voter turnout: Evidence from a large-scale field experiment. *American Political Science Review*, 102(1), 33–48.
- Künzel, S. R., Sekhon, J. S., Bickel, P. J., & Yu, B. (2019). Metalearners for estimating heterogeneous treatment effects using machine learning. *Proceedings of the National Academy of Sciences*, 116(10), 4156–4165.
- Athey, S., Tibshirani, J., & Wager, S. (2019). Generalized random forests. *Annals of Statistics*, 47(2), 1148–1178.
- Holland, P. W. (1986). Statistics and causal inference. *Journal of the American Statistical Association*, 81(396), 945–960.
- Luque-Fernandez, M. A., Schomaker, M., Rachet, B., & Schnitzer, M. E. (2018). Targeted maximum likelihood estimation for a binary treatment: A tutorial. *Statistics in Medicine*, 37(16), 2530–2546.
- Stata Blog (2025-10-22). Heterogeneous treatment effect estimation with S-, T-, and X-learners using h2oml.
